SPP-Perc: Peceptron Learning Based Enhanced Signature Path Prefetcher

### 2018-11-18 abs.tex

\begin{abstract}
  \textbf{Hardware prefetching has been introduced in modern
    processors as a way to hide cache latencies.  An efficient
    prefetcher should be able to identify complex memory access
    patterns during a program execution.  This enables it to fetch a
    block ahead of its demand access, potentially saving a cache miss.
  }

  \textbf{In this paper, we introduce perceptron learning to help make
    this prefetching decision efficiently.  We have taken a state of
    the art prefetcher - Signature Path Prefetcher (SPP) as the
    underlying prefetch engine.  The perceptron layer acts as a
    throttler to filter out the unnecessary prefetches recommended by
    SPP.  We have also explored a range of features that can be used
    to train the perceptron layer.  Our results show that SPP-Perc
    improves performance on the memory intensive subset of the SPEC
    2017 Benchmark suite by 6.84\% on single-core and by 11.9\% on
    multi-core traces, as compared to just SPP.  We also demonstrate
    that the performance gained from using our efficient throttler
    scales better with increasing number of cores.  }
\end{abstract}

>>>>

\begin{abstract}
Hardware prefetching has been introduced in modern
processors as a way to hide cache latencies.  An efficient
prefetcher should be able to identify complex memory access
patterns during a program execution.  This enables it to fetch a
block ahead of its demand access, potentially saving a cache miss.

In this paper, we introduce perceptron learning to help make
this prefetching decision efficiently. The perceptron layer acts 
as a throttler to filter out the unnecessary prefetches recommended 
by the underlying prefetcher.  We have also explored a range 
of features that can be used to train the perceptron layer.  
Our results show that NATCH
improves performance on the memory intensive subset of the SPEC
2017 Benchmark suite by 6.84\% on single-core and by 11.9\% on
multi-core traces, as compared to state-of-the art prefetcher.  
We also demonstrate that the performance gained from using 
our efficient throttler scales better with increasing number 
of cores. 
\end{abstract}

### 2018-11-18 conc.tex

\section{CONCLUSION}
In this paper, we introduced perceptron based learning for data
prefetching in the form of SPP-Perc.  Perceptron acts an orthogonal
throttler to the underlying prefetch engine.  SPP-Perc improves
performance by over XX\% over the baseline, which corresponds to YY\%
over the next best performing prefetcher.  We believe this paper can
lead to interesting future work like generalizing SPP-Perc to work
with any prefetcher with minimal knowledge of the underlying
prefetcher architecture.

>>>>

\section{CONCLUSION}
\label{Conclusion}
In this paper, we introduced neural network based learning for data
prefetching in the form of NATCH.  Perceptron acts an orthogonal
throttler to the underlying prefetch engine.  NATCH improves
performance by over XX\% over the baseline, which corresponds to YY\%
over the next best performing prefetcher.  
We believe that this is a robust and adaptable technique which can
be used to enhance any existing or upcoming prefetchers.

### 2018-11-18 method.tex

\subsection{Performance Model}
\label{Method-Model}
For testing and comparing NATCH, we used ChampSim\cite{Champsim}
simulator.  ChampSim is an enhanced version of the framework that was
used for the 2nd Data Prefetch Championship (DPC-2).\cite{DPC_2} We
model model a 1-core / 4-core / 8-core Out of Order machine. The
details of the configuration parameters is summarized in Table
\ref{tab:Sim_params}.

@@

The block size was fixed 64KB. Prefetching was only triggered at L2
Cache demand accesses but could be sent to L2 Cache or Last Level
Cache.  No L1 Data level prefetching was done.  LRU replacement policy
was used on all levels of cache hierarchies.  Branch prediction was
done using the perceptron branch predictor\cite{Perc_Branch}.  Page
size was kept as 4KB.  Champsim operates all the prefetchers strictly
in the physical address space.

@@

\subsection{Workloads}
\label{Method-Workloads}
This is the first time that SPEC 2017 benchmark suite\cite{SPEC2017}
has been used to characterize and measure the prefetch performance.
We use all the 20 workloads available in the SPEC 2017 Suite.  Using
SimPoint\cite{SimPoint} methodology, we identified 95 different
program segments of 1 Billion instructions each.

We performed cross-validation of our NATCH model using SPEC
2006\cite{SPEC2006} and CloudSuite\cite{CloudSuite} benchmarks.  For
SPEC 2006, we developed 94 simpoints spread across the 29
applications.  For CloudSuite, we used the traces made available for
the 2nd Cache Replacement Competition (CRC-2)\cite{CRC_2}.  The traces
include 4 multi-core applications with 6 distinct phases per
application.  In total, we used 285 traces representing workloads
across 53 applications.

@@

\subsection{Preferchers Simulated}
\label{Method-Prefetchers}
We compared NATCH against three of the latest state of the art
hardware-only prefetchers: Best Offset Prefetcher (BOP), DRAM Aware - Access Map
Pattern Matching (DA-AMPM)\cite{DA_AMPM} and Signature Path Prefecher
(SPP).  BOP was the winner of 2nd Data Prefetching Championship.
DA-AMPM is the enhanced version of AMPM, modified to account for DRAM
row buffer locality.  SPP has been shown to outperform BOP on SPEC
2006 traces.  For each of these, we compare their speedups taking no
prefetching as the baseline.

@@

\begin{table}[]
    \centering
    \begin{tabular}{|p{1.5cm}|p{1cm}|c|p{1.5cm}|}
    \hline
        \textbf{Structure} &
        \textbf{Entry} &
        \textbf{Components} &
        \textbf{Total} \\
    \hline
        \multirow{5}{1.5cm}{Signature Table} & \multirow{5}{1.5cm}{256} & Valid (1 bit) & \multirow{5}{1cm}{11008  bits}\\
        & & Tag (16 bits) &\\
        & & Last Offset (6 bits) &\\  
        & & Signature (12 bits) &\\
        & & LRU (6 bits) &\\
    \hline
        \multirow{3}{1.5cm}{Pattern Table} & \multirow{3}{1.5cm}{512} & $C_{sig}$ (4bits) &\multirow{3}{1cm}{24576 bits}\\
        & & $C_{delta}$ (4*4 bits) &\\
        & & Delta (4*7 bits) &\\
    \hline
        \multirow{3}{1.5cm}{Perceptron Tables} & \multirow{3}{1.5cm}{4096*8 128*1} & \multirow{3}{1.5cm}{5 bits} & \multirow{3}{1.5cm}{164480 bits}\\
        & & &\\
        & & &\\
    \hline
        \multirow{2}{1.5cm}{Prefetch Filter \footnotemark[1]} & \multirow{2}{1.5cm}{1024} & \multirow{2}{1.5cm}{84 bits} & \multirow{2}{1.5cm}{86016 bits}\\
        & & &\\
    \hline
     \multirow{2}{1.5cm}{Reject Filter \footnotemark[2]} & \multirow{2}{1.5cm}{1024} & \multirow{2}{1.5cm}{83 bits} & \multirow{2}{1.5cm}{84992 bits}\\
        & & &\\
    \hline
        \multirow{4}{1.5cm}{Global History Register} & \multirow{4}{1.5cm}{8} & Signature (12 bits) & \multirow{4}{1.5cm}{264 bits}\\
        & & Confidence (8 bits) & \\
        & & Last Offset (6 bits) & \\
        & & Delta (7 bits) & \\
    \hline
        \multirow{2}{1.5cm}{Accuracy Counters} & 1 & C$_{total}$ & 10 bits\\
        & 1 & C$_{useful}$ & 10 bits\\
    \hline
        \multirow{3}{1.5cm}{Global PC Trackers} & \multirow{3}{1cm}{3} & $PC_1$ (12 bits)& \multirow{3}{1.cm}{36 bits}\\
        & & $PC_2$ (12 bits) &\\
        & & $PC_3$ (12 bits) &\\
    \hline
        \multicolumn{4}{|c|}{\textbf{Total: 371356 bits = 45.33 KB}}\\
    \hline
    \end{tabular}
    \caption{NATCH Storage Overhead}
    \label{tab:NATCH_overhead}
\end{table}
\footnotetext[1]{Components of Prefetch Filter can be found in Table \ref{tab:PF_metadata}}
\footnotetext[2]{RF does not need to maintain the useful bit as that only applies for prefetches that ultimately made through}

>>>>

\subsection{Performance Model}
\label{Method-Model}
For testing and comparing NATCH, we used ChampSim~\cite{Champsim}
simulator.  ChampSim is an enhanced version of the framework that was
used for the 2nd Data Prefetch Championship (DPC-2)~\cite{DPC_2}. We
model model a 1-core / 4-core / 8-core Out of Order machine. The
details of the configuration parameters is summarized in Table
\ref{tab:Sim_params}.

@@

The block size was fixed 64KB. Prefetching was only triggered at L2
Cache demand accesses but could be sent to L2 Cache or Last Level
Cache.  No L1 Data level prefetching was done.  LRU replacement policy
was used on all levels of cache hierarchies.  Branch prediction was
done using the perceptron branch predictor~\cite{Perc_Branch}.  Page
size was kept as 4KB.  Champsim operates all the prefetchers strictly
in the physical address space.

@@

\subsection{Workloads}
\label{Method-Workloads}
This is the first time that SPEC 2017 benchmark suite~\cite{SPEC2017}
has been used to characterize and measure the prefetch performance.
We use all the 20 workloads available in the SPEC 2017 Suite.  Using
SimPoint~\cite{SimPoint} methodology, we identified 95 different
program segments of 1 Billion instructions each.

We performed cross-validation of our NATCH model using SPEC
2006~\cite{SPEC2006} and CloudSuite~\cite{CloudSuite} benchmarks.  For
SPEC 2006, we developed 94 simpoints spread across the 29
applications.  For CloudSuite, we used the traces made available for
the 2nd Cache Replacement Competition (CRC-2)~\cite{CRC_2}.  The traces
include 4 multi-core applications with 6 distinct phases per
application.  In total, we used 285 traces representing workloads
across 53 applications.

@@

\subsection{Preferchers Simulated}
\label{Method-Prefetchers}
We compared NATCH against three of the latest state of the art
hardware-only prefetchers: Best Offset Prefetcher (BOP), DRAM Aware - Access Map
Pattern Matching (DA-AMPM)~\cite{DA_AMPM} and Signature Path Prefecher
(SPP).  BOP was the winner of 2nd Data Prefetching Championship.
DA-AMPM is the enhanced version of AMPM, modified to account for DRAM
row buffer locality.  SPP has been shown to outperform BOP on SPEC
2006 traces.  For each of these, we compare their speedups taking no
prefetching as the baseline.

@@

\begin{table}[]
    \centering
    \begin{tabular}{|c|c|c|c|}
    \hline
        \textbf{Structure} &
        \textbf{Entry} &
        \textbf{Components} &
        \textbf{Total} \\
    \hline
        \multirow{5}{2.2cm}{Signature Table} &      & Valid (1 bit)        &             \\
                                             &      & Tag (16 bits)        &             \\
                                             & 256  & Last Offset (6 bits) & 11008 bits  \\  
                                             &      & Signature (12 bits)  &             \\
                                             &      & LRU (6 bits)         &             \\
    \hline
        \multirow{3}{2.2cm}{~~Pattern Table} &      & $C_{sig}$ (4bits)      &               \\
                                             & 512  & $C_{delta}$ (4*4 bits) & 24576 bits    \\
                                             &      & Delta (4*7 bits)       &               \\
    \hline
        \multirow{3}{1.5cm}{Perceptron\newline}     & 4096*8    &            &              \\
        \multirow{2}{0.9cm}{Tables}                 & 2048*x    & 5 bits     & 164480 bits  \\
                                                    & 128*1     &            &              \\
    \hline
        Prefetch Filter\footnotemark[1]             & 1024      & 84 bits    & 86016 bits   \\
    \hline
        Reject Filter\footnotemark[2]               & 1024      & 83 bits    & 84992 bits   \\
    \hline
        \multirow{4}{1cm}{Global\newline\newline}   & \multirow{4}{0.2cm}{8} & Signature (12 bits)  & \multirow{4}{1.1cm}{264 bits} \\
        \multirow{3}{1.1cm}{History\newline}        &                        & Confidence (8 bits)  &                               \\
        \multirow{2}{1.2cm}{Register}               &                        & Last Offset (6 bits) &                               \\
                                                    &                        & Delta (7 bits)       &                               \\
    \hline
        Accuracy        & 1     & C$_{total}$       & 10 bits   \\
        Counters        & 1     & C$_{useful}$      & 10 bits   \\
    \hline
        \multirow{3}{1.5cm}{Global PC\newline}      &       & $PC_1$ (12 bits)      &           \\
        \multirow{2}{1.5cm}{~Trackers}              & 3     & $PC_2$ (12 bits)      & 36 bits   \\
                                                    &       & $PC_3$ (12 bits)      &           \\
    \hline
        \multicolumn{4}{|c|}{\textbf{Total: 371392 bits = 45.33 KB}}\\
    \hline
    \end{tabular}
    \caption{SPP-Perc Storage Overhead}
    \label{tab:SPPPerc_overhead}
\end{table}

\footnotetext[1]{Components of Prefetch Filter can be found in Table \ref{tab:PF_metadata}}
\footnotetext[2]{RF does not need to maintain the useful bit as that only applies for prefetches that ultimately made through}

### 2018-11-18 results.tex

\section{RESULTS}
This section discusses the results obtained from running SPP-Perc, in
terms of cache Misses Per Kilo Instructions (MPKI) and Instruction Per
Cycle (IPC) metrics.  For SPEC 2017 suite, first we present the
results for single-threaded workloads then the results for multi-core
workloads are presented.

\subsection{Single-core Results}
Figure \_\_ shows the IPC speedup of the 4 prefetchers simulated.  All
the results have been normalized to baseline \textit{i.e.}, no
prefetching.  As can be seen, SPP-Perc yields a speedup of XX\% over
the baseline.  This is equivalent to XX\% over DA-AMPM, xx\% over BOP
and XX\% over SPP.  Out of the XX benchmarks, SPP-Perc nearly matches
or outperforms most of the prefetchers on XX traces.

At its peak, SPP-Perc manages to get the IPC speedup of a factor of
over \textbf{3.4} on the trace \textit{602.gcc\_s-2226B}.  This also
corresponds to speedup gain of \textbf{22.4\%} over the next best
prefetcher - SPP.  In general, benchmarks \textit{602.gcc\_s,
  603.bwaves\_s, 605.mcf\_s, 621.wrf\_s} and \textit{649.fotonik3d\_s}
benefit the most from SPP-Perc, as compared to SPP only prefetching.

@@

\subsection{Multi-core Results}
In this section, we demonstrate the improvement achieved by SPP-Perc
for a mix of multi-programmed workloads.

\textit{4-core environment} Figure \_\_ shows a comparison of speedups
obtained on workload mix of memory intensive subset of SPEC 2017.  We
plot all the 4 prefetchers, normalized to baseline.  The workloads
have been sorted in the increasing order of performance benefit, as
measured in terms of normalized weighted speedup.  As can be seen,
SPP-Perc offers an IPC improvement by 33.5\% on these traces.  This is
5.5\% over the next best prefetcher SPP; and XX\% and YY\% over BOP
and DA-AMPM respectively.

This increased improvement by SPP-Perc on a 4-core machine is not
surprising.  Since SPP-Perc has an extremely intelligent throttler
working to eliminate useless prefetches before they can create cache
pollution.  This further amplifies the edge that SPP had over BOP.
With the advantage of an orthogonal prefetcher, SPP can be as
aggressive as possible and leave the accuracy aspect to perceptron
throttler.

\textit{8-core environment} To test the effectiveness of SPP-Perc in a
tighter constrained multi-core environment, we simulated an 8-core
machine with 16 MB last Level Cache.  Other parameters were kept same
as discussed under Methodology in Section XX.  <See if we can use
speed-up over BOP as a measure here>

\subsection{Additional Memory Constraints}
Figure \_\_ shows the performance of SPP-Perc against the other
prefetchers in tighter constraint environments.  For sake of
simplicity, we show 2 traces where SPP-Perc gives the most advantage
over baseline, 2 traces where SPP-Perc doesn't perform as well as the
best prefetcher *<-replace with the prefetcher name once results are
available* and finally GeoMean across all the traces.  <Hopefully BOP
would degrade more here as it is super aggressive.  By definition,
SPP-Perc should be better than SPP as the throttler would better
control the aggressiveness.  In SPP paper DA-AMPM performs close
enough to BOP. That should get replicated here too.>

\subsection{Cross Validation}
Here we try to demonstrate the robustness of SPP-Perc by testing it on
completely unseen benchmarks, namely SPEC 2006 and CloudSuite.

Fig \_\_ shows the speed-up achieved by BOP, SPP and SPP-Perc on
selected traces of SPEC 2006, along with the overall geometric mean
over whole of the benchmark and over memory-intensive subset of the
SPEC 2006 benchmark.  As can be seen, SPP-Perc peovides a speed-up of
44.4\% over baseline, on the memory intensive subset of SPEC 2006
benchmark.  This corresponds to 6.84\% over SPP and XX\% over BOP.  On
the whole of the SPEC 2006 suite, the speedup obtained is 22.4\%,
which is 3.34\% ahead of SPP.

>>>>

\section{RESULTS}
\label{Results}

This section discusses the results obtained from running NATCH, in
terms of cache Misses Per Kilo Instructions (MPKI) and Instruction Per
Cycle (IPC) metrics.  For SPEC 2017 suite, first we present the
results for single-threaded workloads then the results for multi-core
workloads are presented.

\subsection{Single-core Results}
\label{Results-Single}
Figure \_\_ shows the IPC speedup of the 4 prefetchers simulated.  All
the results have been normalized to baseline \textit{i.e.}, no
prefetching.  As can be seen, NATCH yields a speedup of XX\% over
the baseline.  This is equivalent to XX\% over DA-AMPM, xx\% over BOP
and XX\% over SPP.  Out of the XX benchmarks, NATCH nearly matches
or outperforms most of the prefetchers on XX traces.

At its peak, NATCH manages to get the IPC speedup of a factor of
over \textbf{3.4} on the trace \textit{602.gcc\_s-2226B}.  This also
corresponds to speedup gain of \textbf{22.4\%} over the next best
prefetcher - SPP.  In general, benchmarks \textit{602.gcc\_s,
  603.bwaves\_s, 605.mcf\_s, 621.wrf\_s} and \textit{649.fotonik3d\_s}
benefit the most from NATCH, as compared to SPP only prefetching.

@@

\subsection{Multi-core Results}
\label{Results-Multi}
In this section, we demonstrate the improvement achieved by NATCH
for a mix of multi-programmed workloads.

\textit{4-core environment} Figure \_\_ shows a comparison of speedups
obtained on workload mix of memory intensive subset of SPEC 2017.  We
plot all the 4 prefetchers, normalized to baseline.  The workloads
have been sorted in the increasing order of performance benefit, as
measured in terms of normalized weighted speedup.  As can be seen,
NATCH offers an IPC improvement by 33.5\% on these traces.  This is
5.5\% over the next best prefetcher SPP; and XX\% and YY\% over BOP
and DA-AMPM respectively.

This increased improvement by NATCH on a 4-core machine is not
surprising.  Since NATCH has an extremely intelligent throttler
working to eliminate useless prefetches before they can create cache
pollution.  This further amplifies the edge that SPP had over BOP.
With the advantage of an orthogonal prefetcher, SPP can be as
aggressive as possible and leave the accuracy aspect to perceptron
throttler.

\textit{8-core environment} To test the effectiveness of NATCH in a
tighter constrained multi-core environment, we simulated an 8-core
machine with 16 MB last Level Cache.  Other parameters were kept same
as discussed under Methodology in Section \ref{Method-Workloads}.  
\textit{<See if we can use
speed-up over BOP as a measure here>}

\subsection{Additional Memory Constraints}
\label{Results-AdditionalMem}
Figure \_\_ shows the performance of NATCH against the other
prefetchers in tighter constraint environments.  For sake of
simplicity, we show 2 traces where NATCH gives the most advantage
over baseline, 2 traces where NATCH doesn't perform as well as the
best prefetcher *<-replace with the prefetcher name once results are
available* and finally GeoMean across all the traces.

\subsection{Cross Validation}
\label{Results-CrossVal}
Here we try to demonstrate the robustness of NATCH by testing it on
completely unseen benchmarks, namely SPEC 2006 and CloudSuite.

Fig \_\_ shows the speed-up achieved by BOP, SPP and NATCH on
selected traces of SPEC 2006, along with the overall geometric mean
over whole of the benchmark and over memory-intensive subset of the
SPEC 2006 benchmark.  As can be seen, NATCH provides a speed-up of
44.4\% over baseline, on the memory intensive subset of SPEC 2006
benchmark.  This corresponds to 6.84\% over SPP and XX\% over BOP.  On
the whole of the SPEC 2006 suite, the speedup obtained is 22.4\%,
which is 3.34\% ahead of SPP.
